---
title: "Bathing Water Quality Monitor"
format:
  dashboard:
    theme: cosmo
    scrolling: true
    nav-buttons:
      - icon: github
        href: "#"
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(lubridate)
library(plotly)
library(readr)
library(DT)
library(brms)
library(tibble)

# ── Load predictions and raw data ─────────────────────────────────────────────
preds <- read_csv("output/predictions.csv", show_col_types = FALSE) |>
  mutate(year = as.character(year(date)))

rain_raw <- read_csv("data/sademaarat.csv", show_col_types = FALSE) |>
  rename(year_n = Vuosi, month = Kk, day = Pv, rain = `Sademäärä (mm)`) |>
  mutate(date = make_date(year_n, month, day),
         rain = if_else(rain < 0, NA_real_, rain)) |>
  select(date, rain)

flow_raw <- read_delim("data/virtaama.csv", delim = ";",
                       show_col_types = FALSE, trim_ws = TRUE) |>
  rename(date = `Päivä`, flow = virtaama) |>
  mutate(date = dmy(date)) |>
  select(date, flow)

ENTERO_THRESHOLD <- 400
COLI_THRESHOLD   <- 500

entero <- filter(preds, bacteria == "entero")
coli   <- filter(preds, bacteria == "coli")

# Latest observed values for value boxes
latest_entero <- entero |> filter(is_observed) |> slice_max(date)
latest_coli   <- coli   |> filter(is_observed) |> slice_max(date)

# Helper: traffic-light status
status_label <- function(val, threshold) {
  if (is.na(val)) return(list(text = "No data", color = "secondary"))
  if (val > threshold) list(text = "Above limit", color = "danger")
  else list(text = "Below limit", color = "success")
}

entero_status <- status_label(latest_entero$observed, ENTERO_THRESHOLD)
coli_status   <- status_label(latest_coli$observed,   COLI_THRESHOLD)

## Rangebreaks that hide the non-summer months (Sep–May) on date axes, so
## consecutive June-August seasons sit next to each other with no dead space.
SUMMER_RANGEBREAKS <- list(list(pattern = "day of year", bounds = list(244, 151)))

make_bacteria_plot <- function(df, threshold, title_text) {
  obs <- filter(df, is_observed)

  plot_ly() |>
    # CI ribbon
    add_ribbons(data = df, x = ~date,
                ymin = ~pred_lo90, ymax = ~pred_hi90,
                fillcolor = "rgba(70,130,180,0.2)",
                line = list(color = "transparent"),
                name = "90% CI", hoverinfo = "skip") |>
    # Prediction line
    add_lines(data = df, x = ~date, y = ~pred_mean,
              line = list(color = "steelblue", width = 2),
              name = "Model prediction") |>
    # Observed points
    add_markers(data = obs, x = ~date, y = ~observed,
                marker = list(color = "black", size = 7),
                name = "Observed",
                text  = ~paste0(format(date, "%d %b %Y"),
                                "<br>Measured: ", observed, " CFU/100 mL"),
                hovertemplate = "%{text}<extra></extra>") |>
    # Threshold line
    add_lines(x = range(df$date), y = c(threshold, threshold),
              line = list(color = "red", dash = "dash", width = 1.5),
              name = paste("Limit:", threshold),
              hoverinfo = "skip") |>
    layout(
      yaxis  = list(title = "CFU / 100 mL", type = "log",
                    tickformat = ","),
      xaxis  = list(title = "", rangebreaks = SUMMER_RANGEBREAKS),
      legend = list(orientation = "h", y = -0.15),
      hovermode = "x unified"
    ) |>
    config(displayModeBar = FALSE)
}
```

# Overview

## Row {height=15%}

```{r}
#| content: valuebox
#| title: "Latest Entero measurement"
list(
  value = if (nrow(latest_entero) > 0) paste0(latest_entero$observed, " CFU/100 mL") else "—",
  caption = if (nrow(latest_entero) > 0) format(latest_entero$date, "%d %b %Y") else "",
  color = entero_status$color,
  icon  = "droplet"
)
```

```{r}
#| content: valuebox
#| title: "Entero status"
list(
  value   = entero_status$text,
  caption = paste("Limit:", ENTERO_THRESHOLD, "CFU/100 mL"),
  color   = entero_status$color,
  icon    = if (entero_status$color == "success") "check-circle" else "exclamation-triangle"
)
```

```{r}
#| content: valuebox
#| title: "Latest E. coli measurement"
list(
  value   = if (nrow(latest_coli) > 0) paste0(latest_coli$observed, " CFU/100 mL") else "—",
  caption = if (nrow(latest_coli) > 0) format(latest_coli$date, "%d %b %Y") else "",
  color   = coli_status$color,
  icon    = "droplet-fill"
)
```

```{r}
#| content: valuebox
#| title: "E. coli status"
list(
  value   = coli_status$text,
  caption = paste("Limit:", COLI_THRESHOLD, "CFU/100 mL"),
  color   = coli_status$color,
  icon    = if (coli_status$color == "success") "check-circle" else "exclamation-triangle"
)
```

## Row {height=42%}

### Enterococci {.tabset}

```{r}
#| title: "2026"
make_bacteria_plot(filter(entero, year == "2026"), ENTERO_THRESHOLD, "Enterococci 2026")
```

```{r}
#| title: "All years"
make_bacteria_plot(entero, ENTERO_THRESHOLD, "Enterococci")
```

```{r}
#| title: "2020"
make_bacteria_plot(filter(entero, year == "2020"), ENTERO_THRESHOLD, "Enterococci 2020")
```

```{r}
#| title: "2021"
make_bacteria_plot(filter(entero, year == "2021"), ENTERO_THRESHOLD, "Enterococci 2021")
```

```{r}
#| title: "2022"
make_bacteria_plot(filter(entero, year == "2022"), ENTERO_THRESHOLD, "Enterococci 2022")
```

```{r}
#| title: "2023"
make_bacteria_plot(filter(entero, year == "2023"), ENTERO_THRESHOLD, "Enterococci 2023")
```

```{r}
#| title: "2024"
make_bacteria_plot(filter(entero, year == "2024"), ENTERO_THRESHOLD, "Enterococci 2024")
```

```{r}
#| title: "2025"
make_bacteria_plot(filter(entero, year == "2025"), ENTERO_THRESHOLD, "Enterococci 2025")
```

### E. coli {.tabset}

```{r}
#| title: "2026"
make_bacteria_plot(filter(coli, year == "2026"), COLI_THRESHOLD, "E. coli 2026")
```

```{r}
#| title: "All years"
make_bacteria_plot(coli, COLI_THRESHOLD, "E. coli")
```

```{r}
#| title: "2020"
make_bacteria_plot(filter(coli, year == "2020"), COLI_THRESHOLD, "E. coli 2020")
```

```{r}
#| title: "2021"
make_bacteria_plot(filter(coli, year == "2021"), COLI_THRESHOLD, "E. coli 2021")
```

```{r}
#| title: "2022"
make_bacteria_plot(filter(coli, year == "2022"), COLI_THRESHOLD, "E. coli 2022")
```

```{r}
#| title: "2023"
make_bacteria_plot(filter(coli, year == "2023"), COLI_THRESHOLD, "E. coli 2023")
```

```{r}
#| title: "2024"
make_bacteria_plot(filter(coli, year == "2024"), COLI_THRESHOLD, "E. coli 2024")
```

```{r}
#| title: "2025"
make_bacteria_plot(filter(coli, year == "2025"), COLI_THRESHOLD, "E. coli 2025")
```

## Row {height=28%}

### Daily rainfall (mm) {width=50%}

```{r}
# Restricted to the June–August swimming season, matching the model's training data.
rain_plot_data <- rain_raw |>
  filter(date >= min(preds$date) - 21, date <= max(preds$date), month(date) %in% 6:8)

plot_ly(rain_plot_data, x = ~date, y = ~rain, type = "bar",
        marker = list(color = "rgba(70,130,200,0.7)"),
        name = "Rain (mm)",
        hovertemplate = "%{x|%d %b %Y}: %{y:.1f} mm<extra></extra>") |>
  layout(
    xaxis = list(title = "", rangebreaks = SUMMER_RANGEBREAKS),
    yaxis = list(title = "mm"),
    bargap = 0
  ) |>
  config(displayModeBar = FALSE)
```

### River flow (m³/s) {width=50%}

```{r}
flow_plot_data <- flow_raw |>
  filter(date >= min(preds$date) - 21, date <= max(preds$date), month(date) %in% 6:8)

plot_ly(flow_plot_data, x = ~date, y = ~flow,
        type = "scatter", mode = "lines",
        line = list(color = "rgba(0,100,160,0.8)", width = 1.5),
        name = "Flow",
        hovertemplate = "%{x|%d %b %Y}: %{y:.1f} m³/s<extra></extra>") |>
  layout(
    xaxis = list(title = "", rangebreaks = SUMMER_RANGEBREAKS),
    yaxis = list(title = "m³/s")
  ) |>
  config(displayModeBar = FALSE)
```

# Model

## Row {height=15%}

::: {.card}
Bacteria levels are predicted with a Bayesian regression (Student-t likelihood, fit
with brms/Stan), trained only on June–August measurements:

`log(bacteria) ~ rain_decay + log_flow + s(day_of_year) + (1 | year)`

- **rain_decay** — trailing rainfall, exponentially weighted (3-day half-life),
  capturing short-lived combined-sewer-overflow pulses after rain.
- **log_flow** — same-day Vantaanjoki river flow at Oulunkylä, capturing
  watershed-wide wetness and dilution.
- **s(day_of_year)** — a smooth within-season trend (e.g. warmer water → more growth).
- **(1 | year)** — a per-year adjustment for variation not explained by flow/rain.
:::

## Row {height=42%}

```{r}
model_entero <- readRDS("output/model_entero.rds")
model_coli   <- readRDS("output/model_coli.rds")

coef_table <- function(fit, label) {
  as.data.frame(brms::fixef(fit)) |>
    tibble::rownames_to_column("Term") |>
    filter(Term != "Intercept") |>
    transmute(
      Bacteria  = label,
      Term,
      Estimate  = round(Estimate, 3),
      `95% CI`  = paste0("[", round(Q2.5, 3), ", ", round(Q97.5, 3), "]"),
      `Credible effect` = ifelse(Q2.5 > 0 | Q97.5 < 0, "Yes", "No")
    )
}

bind_rows(
  coef_table(model_entero, "Enterococci"),
  coef_table(model_coli,   "E. coli")
) |>
  datatable(options = list(dom = "t", pageLength = 20), rownames = FALSE) |>
  formatStyle(
    "Credible effect",
    backgroundColor = styleEqual(c("Yes", "No"), c("#e0ffe0", "#f0f0f0"))
  )
```

# Measurements

## Row

```{r}
preds |>
  filter(is_observed) |>
  transmute(
    Date        = format(date, "%d %b %Y"),
    Bacteria    = toupper(bacteria),
    `Observed (CFU/100 mL)` = observed,
    `Model prediction`      = round(pred_mean),
    `90% CI low`            = round(pred_lo90),
    `90% CI high`           = round(pred_hi90),
    `Above limit`           = case_when(
      bacteria == "entero" ~ observed > ENTERO_THRESHOLD,
      bacteria == "coli"   ~ observed > COLI_THRESHOLD
    )
  ) |>
  arrange(desc(Date), Bacteria) |>
  datatable(
    options  = list(pageLength = 15, dom = "ftp"),
    rownames = FALSE,
    filter   = "top"
  ) |>
  formatStyle(
    "Above limit",
    backgroundColor = styleEqual(c(TRUE, FALSE),
                                 c("#ffe0e0", "#e0ffe0"))
  )
```
